Assessing Population Density and Crime Patterns in Toronto Neighborhoods (2014–2023)

### 2024-09-27 paper/paper.qmd
Revise small typos.

---
title: "Assessing Population Density and Crime Patterns in Toronto Neighborhoods (2014–2023)"
author: 
  - Yongqi Liu
thanks: "Code and data are available at:https://github.com/Cassieliu77/Crime-Frequency-and-Population-Density.git"
date: 27 September 2024
date-format: long
toc: true 
abstract: "This study examines the relationship between population density and crime distribution across Toronto neighborhoods from 2014 to 2023. The analysis reveals that high-density areas experience higher crime rates, particularly for auto theft and assault, with socioeconomic factors further influencing crime patterns. Based on these findings, we propose targeted interventions, such as improved public surveillance and increased community engagement, to address crime in high-risk areas. This research offers valuable insights for policymakers and urban planners to enhance crime prevention strategies in Toronto."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The average crime rates represent the mean crime rate across 158 neighborhoods from 2014 to 2023, and were calculated to analyze long-term trends and highlight crime types that have seen significant increases. The variables *Crime Type* and *Crime Count* are based on 2023 data (@tbl-summary1).Each neighborhood’s average rank was determined by ranking all 158 neighborhoods based on their total annual crime count. A rank is assigned for each year, and the average rank over the past decade (2014–2023) is calculated by averaging these yearly ranks. A lower rank indicates a higher crime count. Summaries of the top 10 and bottom 10 ranked neighborhoods are provided in @tbl-summary2 and @tbl-summary4.

@@

```{r}
#| label: tbl-summary2
#| tbl-cap: Summary Table for Top10 Crime-Prone Neighborhoods (2014-2023)
#| echo: false
#| warning: false
#| message: false

@@

\newpage
# Results
## Crime Rates Over the Past Decade
@fig-rates uses the variable *Average Crime Rate* (@tbl-summary3) to show the crime rate trend between 2014 and 2023 and demonstrate the evolution of different crime types in the city of Toronto. It can be seen that the visualization reveals certain patterns in some specific crime types.

@@

@fig-rates reveals a significant upward trend in **Auto Theft** and **Assault**, particularly after 2020. The crime rate for **Assault** consistently remained the highest across all crime types, starting at around 600 per 100,000 population in 2014 and reaching a peak of approximately 800 per 100,000 population in 2023. This consistent rise suggests that assault has been a persistent issue in Toronto, with an even sharper increase following the onset of the COVID-19 pandemic.**Break and Enter**,**Robbery** and **Theft Over** all experienced increases after 2021. This rise suggests a sharp increase in private property related crimes, possibly influenced by socioeconomic factors or changes in law enforcement or technology. In contrast, **Homicide** and **Shooting** had relatively lower crime rates and displayed more stability over time, with little significant change compared to **Assault** and **Auto Theft**.

@@

crime_data_sf <- readRDS(here::here("data/analysis_data/cleaning_crime_data.rds"))
ggplot(crime_data_sf) +
  geom_sf(aes(fill = POPULATION_2023), color = "black", alpha = 0.9) +
  scale_fill_gradient(low = "#d3e5ff", high = "#08306b", name = "Population") +
  geom_text(aes(label = HOOD_ID, geometry = geometry),
    stat = "sf_coordinates", size = 4, color = "white", check_overlap = TRUE
  ) +
  labs(
    title = "Toronto Neighbourhood Population Map",
    subtitle = "Based on 2023 Population",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```
\newpage

@@

```{r}
#| label: fig-rank-crime
#| fig-cap: Top10 Ranked Crime-Prone Neigborhoods (2014-2023)
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: tbl-summary4
#| tbl-cap: Top10 Ranked Safest Neigborhoods (2014-2023)
#| echo: false
#| warning: false
#| message: false

>>>>

---
title: "Assessing Population Density and Crime Patterns in Toronto Neighborhoods (2014–2023)"
author: 
  - Yongqi Liu
thanks: "Code and data are available at:https://github.com/Cassieliu77/Crime-Frequency-and-Population-Density.git"
date: 27 September 2024
date-format: long
toc: true 
abstract: "This paper examines the relationship between population density and crime distribution across Toronto neighborhoods from 2014 to 2023. The analysis reveals that high-density areas experience higher crime rates, particularly for auto theft and assault, with socioeconomic factors further influencing crime patterns. Based on these findings, we propose targeted interventions and offers data-driven suggestions for policymakers and urban planners to enhance crime prevention strategies in Toronto."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The average crime rates represent the mean crime rate across 158 neighborhoods from 2014 to 2023, and were calculated to analyze long-term trends and highlight crime types that have seen significant increases. The variables *Crime Type* and *Crime Count* are based on 2023 data (@tbl-summary1).Each neighborhood’s average rank was determined by ranking all 158 neighborhoods based on their total annual crime count. A rank is assigned for each year, and the average rank over the past decade (2014–2023) is calculated by averaging these yearly ranks. A lower rank indicates a higher crime count. Summaries of the Top 10 and Bottom 10 ranked neighborhoods are provided in @tbl-summary2 and @tbl-summary4.

@@

```{r}
#| label: tbl-summary2
#| tbl-cap: Summary Table for Top 10 Crime-Prone Neighborhoods (2014-2023)
#| echo: false
#| warning: false
#| message: false

@@

# Results
## Crime Rates Over the Past Decade
@fig-rates uses the variable *Average Crime Rate* (@tbl-summary3) to show the crime rate trend between 2014 and 2023 and demonstrate the evolution of different crime types in the city of Toronto. It can be seen that the visualization reveals certain patterns in some specific crime types.

@@

@fig-rates reveals a significant upward trend in **Auto Theft** and **Assault**, particularly after 2020. The crime rate for **Assault** consistently remained the highest across all crime types, starting at around 600 per 100,000 population in 2014 and reaching a peak of approximately 800 per 100,000 population in 2023. This consistent rise suggests that assault has been a persistent issue in Toronto, with an even sharper increase following the onset of the COVID-19 pandemic. **Break and Enter**,**Robbery** and **Theft Over** all experienced increases after 2021. This rise suggests a sharp increase in private property related crimes, possibly influenced by socioeconomic factors or changes in law enforcement or technology. In contrast, **Homicide** and **Shooting** had relatively lower crime rates and displayed more stability over time, with little significant change compared to **Assault** and **Auto Theft**.

@@

crime_data_sf <- readRDS(here::here("data/analysis_data/cleaning_crime_data.rds"))
ggplot(crime_data_sf) +
  geom_sf(aes(fill = POPULATION_2023), color = "black", alpha = 0.9) +
  scale_fill_gradient(low = "#d3e5ff", high = "#08306b", name = "Population") +
  geom_text(aes(label = HOOD_ID, geometry = geometry),
    stat = "sf_coordinates", size = 4, color = "white", check_overlap = TRUE
  ) +
  labs(
    title = "Toronto Neighbourhood Population Map",
    subtitle = "Based on 2023 Population",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 13)
  )
```
\newpage

@@

```{r}
#| label: fig-rank-crime
#| fig-cap: Top 10 Ranked Crime-Prone Neigborhoods (2014-2023)
#| echo: false
#| warning: false
#| message: false

@@

```{r}
#| label: tbl-summary4
#| tbl-cap: Top 10 Ranked Safest Neigborhoods (2014-2023)
#| echo: false
#| warning: false
#| message: false
